Fairly Random: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}

### 2023-09-05 ms/ms.tex
updated paper

\title{Asymmetry in Cricket: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}

\author{
Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
Derek Willis\thanks{Lecturer in Data and Computational Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}; Corresponding author. Phone: 650 269 7946. Address: 156th Place NE, Redmond, WA, 98052.}\;\;
Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
}
\date{\today}

@@

In competitive sports, teams fight for every little advantage. Many sports, however, leave some important things to chance. In cricket, a coin toss is used to decide which team bats first. The choice is widely seen as consequential. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}.} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.

In this paper, we measure the advantage conferred by winning the toss. To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions.

@@

The rest of the paper is organized as follows. We start by describing the data. We then analyze if the data are consistent with the toss being random. Next, we present the results. We end with discussing the policy implications.

@@

\section{Batting Vs. Fielding First}

@@

\subsection{The Difficulty in Estimating the Advantage to Electing to Bat/Field First}

@@

We make these points more concrete. Recall that we saw that 56.5\% of teams that won the toss elect to bat, which is 6.5 percentage points more than what you would expect if the toss had no effect on the batting/fielding decision. In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity. The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sign of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins? No, because these excess wins could have been generated by the defiers making good decisions to field. Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility. Consequently, it is unlikely the LATE can be identified in our setting.

>>>>

\title{Fairly Random: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}

\author{
Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
Derek Willis\thanks{Lecturer in Data and Computational Journalism, Philip Merrill College of Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}.}\;\;
Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
}
\date{\today}

@@

In competitive sports, teams fight for every little advantage. Many sports, however, leave some important things to chance. In cricket, for instance, a coin toss is used to decide which team bats first. The choice is widely seen as consequential. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}.} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.

In this paper, we measure the advantage conferred by winning the toss. To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used before \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but critical pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on decisions made after the toss, e.g., whether the team elects to bat or bowl first. (We tackle this issue at length later in the paper.)

@@

The rest of the paper is organized as follows. We start by describing the data. We then analyze if the data are consistent with the toss being random. Next, we present the results on the effect of winning the toss on the probability of winning. Given that some previous papers condition on decisions made after the toss, we discuss what teams decide to do when they win the toss and why conditioning on post-treatment decisions could bias the inference. We end with discussing the policy implications.

@@

\section{The Decision to Bat or Bowl First}

@@

\subsection{The Difficulty in Estimating the Advantage of Choosing to Bat or Bowl First}

@@

We make these points more concrete. Recall that we saw that 56.5\% of teams that won the toss elect to bat, which is 6.5 percentage points more than what you would expect if the toss had no effect on the batting/fielding decision. In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity. The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sign of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins? No, because these excess wins could have been generated by the defiers making good decisions to field first. Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility. Consequently, it is unlikely the LATE can be identified in our setting.
